---
title: "Prioritizing potential aquaculture"
author: "Aakriti Poudel"
date: "11/13/2025"
format:
  pdf:
    code-fold: false
    toc: true
  html: 
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
editor: source
editor_options: 
  chunk_output_type: inline
---

## Screenshot of GitHub README.md

(place holder)

## Load required libraries

```{r}
library(tidyverse)
library(sf)
library(tmap)
library(terra)
library(kableExtra)
#library(stars)
```

## Read data

```{r}
#| output: false
#| warning: false
#| message: false

# Read sea surface temperature data
sst_2008 <- rast(here::here('data', 'average_annual_sst_2008.tif'))
sst_2009 <- rast(here::here('data', 'average_annual_sst_2009.tif'))
sst_2010 <- rast(here::here('data', 'average_annual_sst_2010.tif'))
sst_2011 <- rast(here::here('data', 'average_annual_sst_2011.tif'))
sst_2012 <- rast(here::here('data', 'average_annual_sst_2012.tif'))

# Combine SST raster into a raster stack
sst_stack <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012,
               along = list(year = 2008:2012))

# Read Bathymetry data
depth <- rast(here::here('data', 'depth.tif'))

# Read Exclusive Economic Zones data
eez <- st_read(here::here('data', 'wc_regions_clean.shp'))

```

## Prepare data

```{r}
# Check CRS for all data set
print(paste0("The CRS for sst_stack is: ", st_crs(sst_stack)$epsg))
print(paste0("The CRS for ocean depth is: ", st_crs(depth)$epsg))
print(paste0("The CRS for eez is: ", st_crs(eez)$epsg))
```

```{r}
# Convert the CRS of sst_stack to EPSG:4326
crs(sst_stack) <- "EPSG:4326" 

# Check CRS of the dataset
if(st_crs(sst_stack) == st_crs(depth)){
print("Coordinate reference systems match!")
} else {
warning("Update coordinate reference systems to match!")
}
```

## Process data

**Find the mean of SST stacks from 2008-2012**

```{r}
# Find the mean SST from 2008-2012 (e.g. create single raster of average SST)
sst_mean <- app(sst_stack, mean, na.rm = TRUE)

# Convert average SST from Kelvin to Celsius
sst_mean_celsius <- sst_mean - 273.15

cat("\nConverted SST range (Celsius):\n")
print(range(sst_mean_celsius[[1]], na.rm = TRUE))
```

**Crop depth raster to match the extent of the SST raster**

```{r}
cat("Ocean depth extent:\n")
print(st_bbox(depth))
cat("\nSST mean celsius extent:\n")
print(st_bbox(sst_mean_celsius))

# Crop the ocean depth 
depth_cropped <- crop(depth, ext(sst_mean_celsius))

```

```{r}
plot(depth_cropped, downsample = 0)
```

**Resample the depth data to match the resolution of the SST data using the nearest neighbor approach**

```{r}
# Resample the ocean depth data
depth_resample <- resample(depth_cropped, sst_mean_celsius, method = 'near')

# Check if they match using compareGeom (better for terra)
compareGeom(depth_resample, sst_mean_celsius, stopOnError = FALSE)

identical(ext(depth_resample), ext(sst_mean_celsius))
```


```{r}
# Method 2: all.equal with tolerance
all.equal(ext(depth_resample), ext(sst_mean_celsius))
```
```{r}
ext(depth_resample) == ext(sst_mean_celsius)
```


```{r}
plot(depth_resample)
```

**Check that the ocean depth and SST match in resolution, extent and coordinate reference system** - Can the rasters be stacked?

```{r}
# Can the rasters be stacked?
if(all.equal(ext(depth_resample), ext(sst_mean_celsius)) == TRUE &&
   all.equal(res(depth_resample), res(sst_mean_celsius)) == TRUE &&
   crs(depth_resample) == crs(sst_mean_celsius)) {
  test_stack <- c(depth_resample, sst_mean_celsius)
  print('Rasters can be successfully stacked!')
} else {
  warning('Rasters cannot be stacked. Compatibility issue detected!')
}
```

## Find suitable locations for Oysters

**Reclassify SST and ocean depth data into locations that are suitable for Oysters**

```{r}
# Reclassify SST into locations that are suitable for oysters
sst_reclass_oysters <- matrix(c(-Inf, 11, 0,
                        11, 30, 1,
                        30, Inf, 0), ncol = 3, byrow = TRUE)

sst_classify_oysters <- classify(sst_mean_celsius, sst_reclass_oysters)

depth_reclass_oysters <- matrix(c(-Inf, -70, 0,
                        -70, 0, 1,
                        0, Inf, 0), ncol = 3, byrow = TRUE)

depth_classify_oysters <- classify(depth_resample, depth_reclass_oysters)

# Find suitable locations 
suitable_locations_oysters <- sst_classify_oysters * depth_classify_oysters

# Check
unsuitable_oysters <- sum(values(suitable_locations_oysters) == 0, na.rm = TRUE)
suitable_oysters <- sum(values(suitable_locations_oysters) == 1, na.rm = TRUE)
total_oysters <- unsuitable_oysters + suitable_oysters

print(paste0("The unsuitable number of location for Oysters is: ", unsuitable_oysters))
print(paste0("The suitable number of location for Oysters is: ", suitable_oysters))
```

```{r}
plot(suitable_locations_oysters)
```

## Choose your own species

***You can find information on species depth and temperature requirements on SeaLifeBase. Remember, we are thinking about the potential for marine aquaculture, so these species should have some reasonable potential for commercial consumption.*** 

My species is **Atlantic Bay Scallop** ***(Argopecten irradians)***. They need the following conditions for optimal growth:
- sea surface temperature: 24 - 26°C
- depth range: 0 - 20 meters below sea level

```{r}
# Reclassify SST into locations that are suitable for Atlantic Bay Scallop
sst_reclass_abs <- matrix(c(-Inf, 24, 0,
                            24, 26, 1,
                            26, Inf, 0), ncol = 3, byrow = TRUE)

sst_classify_abs <- classify(sst_mean_celsius, sst_reclass_abs)

depth_reclass_abs <- matrix(c(-Inf, -20, 0,
                            0, 20, 1,
                            20, Inf, 0), ncol = 3, byrow = TRUE)

depth_classify_abs <- classify(depth_resample, depth_reclass_abs)

# Find suitable locations 
suitable_locations_abs <- sst_classify_abs * depth_classify_abs

# Check
unsuitable_abs <- sum(values(suitable_locations_abs) == 0, na.rm = TRUE)
suitable_abs <- sum(values(suitable_locations_abs) == 1, na.rm = TRUE)
#total_abs <- unsuitable_abs + suitable_abs

print(paste0("The unsuitable number of location for Atlantic Bay Scallop is: ", unsuitable_abs))
print(paste0("The suitable number of location for Atlantic Bay Scallop is: ", suitable_abs))

```

```{r}
plot(suitable_locations_abs)
```
Mytilus californianus
California mussel
depth range usually 0 - 30 m
8.6 - 14.3


```{r}
# Reclassify SST into locations that are suitable for California mussel
sst_reclass_cm <- matrix(c(-Inf, 8.6, 0,
                            8.6, 14.3, 1,
                            14.3, Inf, 0), ncol = 3, byrow = TRUE)

sst_classify_cm <- classify(sst_mean_celsius, sst_reclass_cm)

depth_reclass_cm <- matrix(c(-Inf, -30, 0,
                            0, 30, 1,
                            30, Inf, 0), ncol = 3, byrow = TRUE)

depth_classify_cm <- classify(depth_resample, depth_reclass_cm)

# Find suitable locations 
suitable_locations_cm <- sst_classify_cm * depth_classify_cm

# Check
unsuitable_cm <- sum(values(suitable_locations_cm) == 0, na.rm = TRUE)
suitable_cm <- sum(values(suitable_locations_cm) == 1, na.rm = TRUE)
total_cm <- unsuitable_cm + suitable_cm

print(paste0("The unsuitable number of location for Atlantic Bay Scallop is: ", unsuitable_cm))
print(paste0("The suitable number of location for Atlantic Bay Scallop is: ", suitable_cm))

```

```{r}
plot(suitable_locations_cm)
```

## Determine the most suitable EEZ

**Select suitable cells within West Coast EEZs**

```{r}
#| output: true
# Check and transform coordinate reference systems
if(st_crs(eez) == crs(sst_mean_celsius)) {
  print("Coordinate reference systems match!")
} else {
  warning("Updating coordinate reference systems to match!")
}

# Transform economic zones data set to match raster CRS
eez <- st_transform(eez, crs(sst_mean_celsius))
```

```{r}
# Convert the eez polygons to raster using the regions ID
eez_raster <- rasterize(eez, suitable_locations_oysters, field = "rgn_id")

# Calculate the area in km square
area_km2_cell <- cellSize(suitable_locations_oysters, unit = "km")

# Calculate the suitable area 
suitable_area <- suitable_locations_oysters * area_km2_cell

# Calculate suitable area for each eez polygons
eez_suitable_area <- extract(suitable_area, eez, fun = "sum", na.rm = TRUE)

# Add eez suitable area to the eez dataset
eez$suitable_area_km2 <- eez_suitable_area[, 2]

# Sort by suitable area (descending)
eez_rank <- eez[order(-eez$suitable_area_km2), ]
```

### Visualize the most suitable EEZ

```{r}
#| output: true
# Convert to data frame and select columns
eez_df <- as.data.frame(eez_rank)[, c("rgn", "suitable_area_km2")]

# Visualize the eez rank in table format
kable(eez_df, 
      caption = "<strong style='font-size: 14px; color: #2c3e50;'>Major Oyster Producing Economic Regions in West Coast, US</strong>",
      col.names = c("Region", "Suitable Area (km²)"),
      digits = 2, 
      row.names = FALSE,
      escape = FALSE,
      format.args = list(big.mark = ",")) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left",
    font_size = 13
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "darkolivegreen") %>%
  column_spec(1, bold = TRUE, color = "black", width = "70%") %>%
  column_spec(2, color = "black", width = "50%")
  
```

```{r}
# Plot a map showing the most suitable EEZ
tm_shape(eez_rank) +
  tm_polygons( fill = "suitable_area_km2",
               fill.scale = tm_scale(values = "plasma"),
               fill.legend = tm_legend(title = "Suitable area (per km²)")) + tm_title(text = "West Coast EEZs Ranked by Suitable Area\nfor Oyster Aquaculture") +
  tm_graticules() +
  tm_basemap("Esri.OceanBasemap", alpha = 0.7) +
  tm_layout(inner.margins = c(0.02, 0.02, 0.02, 0.2), component.autoscale = FALSE)
```



